---
title: "Energy Report Code"
author: "Steven Meeks"
date: "2025-11-04"
output: html_document
---

# Load Data

-   CHANGE PATH TO FILE

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(datasets)
library(fpp3)
library(forecast)
library(fable)
library(fabletools)

#Read in the data
energy = read_csv('https://raw.githubusercontent.com/sjsimmo2/TimeSeries/refs/heads/master/energy_F2025.csv')
```

## Aggregate Data

-   Forecasting total daily energy load

```{r}
#aggregate mw by daily total
energy_daily <- energy %>% 
  mutate(date=mdy_hm(datetime_beginning_ept))%>%
  mutate(day = as_date(date)) %>% 
  group_by(day) %>% 
  summarise(daily_total = sum(mw, na.rm = TRUE)) 

#as tsibble
energy.ts<-energy_daily %>%
  as_tsibble(index = day)

energy.ts 
```

## Train/Test Split
```{r}
# Training (Aug 2018-Aug 2023)
train <- energy.ts |> 
  filter(day >= as_date('2018-08-01') & day <= as_date('2023-08-31')) 

# Validation (Sep 2023-Aug 2024)
valid <- energy.ts |> 
  filter(day >= as_date('2023-09-01') & day <= as_date('2024-08-31')) 

# Testing (Sep 2024-Aug 2025)
test <- energy.ts |> 
  filter(day >= as_date('2024-09-01') & day <= as_date('2025-01-31'))
```

## View Time Plot
```{r}
#time plot
autoplot(energy.ts, daily_total)
```

# ESM

## Find Best Model
```{r}
#Fit models
energy_fit <- train |>
  model(
    SES = ETS(daily_total ~ error("A") + trend("N") + season("N")),
    Linear = ETS(daily_total ~ error('A') + trend('A') + season("N")),
    LinearDamp = ETS(daily_total ~ error('A') + trend('Ad') + season("N")),
    HWAdd = ETS(daily_total ~ error("A") + trend("A") + season("A")),
    HWMult = ETS(daily_total ~ error("M") + trend("A") + season("M")),
    Search1 = ETS(daily_total)
)

#goodness of fit
glance(energy_fit) |>
  arrange(AIC)
  #search model - lowest AIC

#FORECASTING
#forecast for all models
energy.for <- energy_fit |> fabletools::forecast(h = 366)

#accuracy
fabletools::accuracy(energy.for, valid) |>
  arrange(MAPE)

#Examine search model
energy_fit |>
  select(Search1) |>
  report()

#Search Model --> MNA components
  #multiplicative error
  #no trend
  #additive seasonality
```

Best Model: Search ETS(M,N,A)

-   Validation Performance
    -   MAPE = 10.23
    -   MAE = 11332.69

## ESM Forecast Plot

```{r}
#ETS search model
energy_esm <- train |> model(ETS(daily_total ~ error("M") + trend("N") + season("A")))
#forecasts
energy.for2 <- energy_esm |> fabletools::forecast(h = 366)
                                    # forecast length of validation

library(ggplot2)

#Plot forecast vs actual
ggplot() +
  geom_line(data = valid, aes(x = day, y = daily_total, colour = "Actual"), linewidth = 0.95) +
  geom_line(data = energy.for2, aes(x = day, y = .mean, colour = "Forecast"), linewidth = 0.95) +
  scale_colour_manual(name = "Data", values = c("Actual" = "black", "Forecast" = "#2596be")) +
  labs(
    title = "ESM Model Validation Performance: Actual vs. Forecast",
    subtitle = "Total Daily Energy Load",
    x = "Month",
    y = "Energy in Megawatts (MW)"
  ) +
  theme_minimal()
```

# ARIMA

Objective: Seasonal ARIMA

-   describe approach for selecting lags
-   forecast model to validation
-   calculate MAE and MAPE for validation

## Seasonal Stationarity

```{r}
#unit root test - weekly period
train %>%
  features(daily_total, unitroot_nsdiffs, period = 7) #weekly period (yearly is too large)
    # no seasonal differences

#STL decomp
dcmp <- train |> 
  model(stl = STL(daily_total)) 

#components
components(dcmp) |> 
  autoplot() + theme_classic()
    #season_year 
    #season_week

# feature strength - weekly
train |>
features(daily_total, feat_stl, period=7)
  # Fs < 0.64 -- no seasonal differencing
```

## Seasonality

### Fourier Terms

```{r}
# Fourier Terms - Weekly and Yearly

# ranges for k
k_weekly  <- 1:3
K_yearly <- 1:13

# grid of combinations
K_grid <- expand.grid(Kw = k_weekly, Ky = K_yearly)

# all fourier term combos
model_defs <- setNames(
  lapply(seq_len(nrow(K_grid)), function(i) {
    ARIMA(
      daily_total ~
        fourier(period=7, K=K_grid$Kw[i]) +
        fourier(period=365.25, K=K_grid$Ky[i]) +
        pdq(0,0,0) + PDQ(0,0,0)
    )
  }),
  paste0("KW", K_grid$Kw, "_KY", K_grid$Ky)
)

# Fit all models
model_FYW <- train |>
  model(!!!model_defs)

# examine goodness of fit
glance(model_FYW) |>
  arrange(AICc)

#Best
  ## Weekly - k=3 is consistently better at same yearly-k
  ## yearly - concern for overfit - favoring largest k
```

### Plot Yearly-K

```{r}
#Plot Yearly K by Info Criteria

#prep data
grid_stats <- model_FYW |> 
  glance() |>
  mutate(
    Kw = as.numeric(sub("KW([0-9]+)_KY([0-9]+)", "\\1", .model)),
    Ky = as.numeric(sub("KW([0-9]+)_KY([0-9]+)", "\\2", .model))
  )

#selected weekly k
chosen_weekly_K <- 3

#format data
elbow_data <- grid_stats |> 
  filter(Kw == chosen_weekly_K) |>
  select(Ky, AICc)

# Reshape for plotting
elbow_long <- elbow_data |>
  pivot_longer(cols = c(AICc), names_to = "Criterion", values_to = "Value")

#plot
ggplot(elbow_long, aes(x = Ky, y = Value)) +
  geom_line(size = 1, color = '#2596be') +
  geom_point(color = '#2596be') +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 5)) +
  theme_minimal() +
  labs(
    title = "Yearly Fourier Term Selection Plot",
    subtitle = "Weekly K = 3",
    x = "Number of Yearly Fourier K Terms",
    y = "Corrected Akaike Information Criteria (AICc) ",
  )

# K=2 limits complexity without sacrificing much AICc
```

## Residuals

### Stationarity
```{r}
#final fourier only model
season_model <- train |> model(ARIMA(daily_total ~ fourier(period=7, K=3) + fourier(period=365.25, K=2) + pdq(0,0,0) + PDQ(0,0,0)))

#stationarity tests
augment(season_model) %>% features(.resid, unitroot_kpss)
augment(season_model) %>% features(.resid, unitroot_ndiffs)
    # ndiffs = 1 difference needed
    # kpss = reject null of stationarity

#Need 1 difference

#add difference
season_model2 <- train |> model(ARIMA(daily_total ~ fourier(period=7, K=3) + fourier(period=365.25, K=2) + pdq(0,1,0) + PDQ(0,0,0)))

#stationarity tests
augment(season_model2) %>% features(.resid, unitroot_kpss)
augment(season_model2) %>% features(.resid, unitroot_ndiffs)

#No further differences
```

### Correlation Plot

```{r}
#overview
season_model2 |>
  gg_tsresiduals(lag=21)

#residuals
model_residuals <- residuals(season_model2)
residuals <- my_residuals$.resid

#ACF Plot
acf_plot <- ggAcf(residuals, lag.max = 21) +
    theme_minimal() +
  labs(
    title = "ACF of Residuals",
    subtitle = paste("Model Residuals ( N =", length(residuals),")"),
    y = "Autocorrelation"
  ) +
  # Customize line and bar appearance for a clean look
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray30")
  )

#PACF Plot
pacf_plot <- ggPacf(residuals, lag.max = 21) +
  # Customize appearance
  theme_minimal() +
  labs(
    title = "PACF of Residuals",
    subtitle = paste("Model Residuals ( N =",length(residuals),")"),
    y = "Partial Autocorrelation"
  ) +
  # Match theme elements
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray30")
  )



print(acf_plot)
print(pacf_plot)


#ACF
  # 4 spikes -- try a few non-seasonal MA
  # spikes 14-15  -- try seasonal MA
#PACF
  # decreasing spikes up to 15 -- try some AR terms
  #try a few seasonal AR
```


## Seasonal ARIMA

### ARMA Models Loop

- Check AR terms ~4 based on spikes in PACF
- Check MA terms ~2 based on spikes in ACF
- Seasonal 1 AR or MA terms

```{r}
# Define search ranges
p_lag <- 3:4
q_lag <- 1:2
P_lag <- 0:2
Q_lag <- 0:2

# Create parameter grid
param_grid <- expand.grid(p = p_lag, d = 1, q = q_lag, P = P_lag, D = 0, Q = Q_lag)


# Fit models
models <- param_grid %>%
  mutate(
    model = pmap(list(p, d, q, P, D, Q), function(p, d, q, P, D, Q) {
      train %>%
        model(
          ARIMA(
            daily_total ~ 0 + fourier(period = 7, K = 3) +
              fourier(period = 365.25, K = 2) +
              pdq(p, d, q) + PDQ(P, D, Q)
          )
        )
    })
  )

# AICc for comparison
results <- models %>%
  mutate(
    AICc = map_dbl(model, ~ glance(.x)$AICc)
  ) %>%
  arrange(AICc)

# View top models
head(results)
```
- AR3MA1_AR2MA2 = 37795.14
- AR4MA1_AR2MA2 = 37796.97
- AR3MA2_AR2MA2 = 37797.34
- AR3MA1_AR1    = 37797.37
- AR3MA1_MA1    = 33797.43


```{r}
#top 5 models by AICc for further comparison
top_models <- train %>%
  model(
  AR3MA1_AR2MA2 = ARIMA(daily_total ~ 0 + fourier(period=7, K=3) + fourier(period=365.25, K=2) + pdq(3,1,1) + PDQ(2,0,2)),
  AR4MA1_AR2MA2 = ARIMA(daily_total ~ 0 + fourier(period=7, K=3) + fourier(period=365.25, K=2) + pdq(4,1,1) + PDQ(2,0,2)),
  AR3MA2_AR2MA2 = ARIMA(daily_total ~ 0 + fourier(period=7, K=3) + fourier(period=365.25, K=2) + pdq(3,1,2) + PDQ(2,0,2)),
  AR3MA1_AR1 = ARIMA(daily_total ~ 0 + fourier(period=7, K=3) + fourier(period=365.25, K=2) + pdq(3,1,1) + PDQ(1,0,0)),
  AR3MA1_MA1 = ARIMA(daily_total ~ 0 + fourier(period=7, K=3) + fourier(period=365.25, K=2) + pdq(3,1,1) + PDQ(0,0,1)),
  #Add search method
  Search = ARIMA(daily_total ~ 0 + fourier(period=7, K=3) + fourier(period=365.25, K=2) + pdq(,1,) + PDQ(,0,))
)
```

### Performance
```{r}
#FORECASTING
#forecast for top models
arima.for <- top_models |> fabletools::forecast(h = 366)

#compare accuracy
fabletools::accuracy(arima.for, valid) |>
  arrange(MAPE)

#view search 
top_models |>
  select(Search) |>
  report()

#search does not produce white noise

#final model
  #SARIMA(3,1,1)(2,0,2) model - lowest MAE and MAPE after search
    # best generalized model - has white noise
```
Best Model: SARIMAX(3,1,1)(2,0,2)

-   Validation Performance
    -   MAPE = 6.806
    -   MAE = 7296.023


### White Noise
```{r}
#test white noise

#search model
search <- train |> model(ARIMA(daily_total ~ 0 + fourier(period=7, K=3) + fourier(period=365.25, K=2) + pdq(0,1,4) + PDQ(1,0,0)))

#View residuals
search |>
  gg_tsresiduals(lag=21)

#white noise?
augment(search) |>
  features(.innov, ljung_box, lag = 21, dof = 5)
  #Search does not have white noise



#hand model
sarimax <- train |> model(ARIMA(daily_total ~ 0 + fourier(period=7, K=3) + fourier(period=365.25, K=2) + pdq(3,1,1) + PDQ(2,0,2)))

#View residuals
sarimax |>
  gg_tsresiduals(lag=21)

#Ljung-box
augment(sarimax) |>
  features(.innov, ljung_box, lag = 21, dof = 8)
  #hand model has white noise
```


## Forecast Plot
```{r}
#forecast
arima.for2 <- sarimax |> fabletools::forecast(h = 365)
                                    # forecast length of validation

fabletools::accuracy(arima.for2, valid) |>
  arrange(MAPE)

library(ggplot2)

#Plot forecast vs actual
ggplot() +
  geom_line(data = valid, aes(x = day, y = daily_total, colour = "Actual"), linewidth = 0.95) +
  geom_line(data = arima.for2, aes(x = day, y = .mean, colour = "Forecast"), linewidth = 0.95) +
  scale_colour_manual(name = "Data", values = c("Actual" = "black", "Forecast" = "#2596be")) +
  labs(
    title = "SARIMA Model Validation Performance: Actual vs. Forecast",
    subtitle = "Total Daily Energy Load",
    x = "Month",
    y = "Energy in Megawatts (MW)"
  ) +
  theme_minimal()
```




